Fast Random Integer Generation in an Interval

### 2023-12-04 Paper/main.tex
fix algo numbering!

\subsubsection{OpenBSD}\label{sec:2.2.1}
We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
The key idea is that this interval does not have to be on the right.
Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
This yields \hyperref[alg:openbsd]{Algorithm 4} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.

@@

\subsubsection{Java}\label{sec:2.2.2}
Since we expect remainder operations to have high latency, always needing to compute two remainders can be quite costly.
Hence, the Java language, in its \texttt{Random class}, uses a different approach.
The key insight is that when we draw a random integer $x \in [0,2^W)$ and compute its remainder $r = x \Mod n$, we can map all integers in $[0,2^W)$ to a multiple of $n$ by computing $x - r$.
Then, every number in the last interval in \hyperref[eq:1]{equation (1)} gets mapped to the last multiple $y$ of $n$ which is not greater than $2^W$.
Note that this implies $y > 2^W - n$.
Every other integer is mapped to a different (and thus smaller) multiple $z$ of $n$ for which $z \leq y - n$ and thus $z \leq 2^W - n$.
Consequently, if we reject $x$ if $x - r > 2^W - n$ and otherwise return $r$, we get a unbiased generator.
The details can be seen in \hyperref[alg:java]{Algorithm 5}.

@@

\subsubsection{Fast-Dice-Roller}\label{sec:2.2.3}
For all prior unbiased algorithms, we need to use at least one remainder operation to debias the algorithm.
Jérémie Lumbroso however devised an algorithm for sampling uniform random integers in $[0,n)$ for a given $n$ by iteratively building up the final number bit-by-bit~\cite{Flips}.
In comparison to previous algorithms, we do not use the whole $W$ bits generated by the PRNG but instead only use one bit at a time building up the final number consisting of $\lceil\log_2n\rceil + 1$ bits.
If we, at some point, overshoot and generate a number bigger than $n$ (but smaller than $2n$) we take a step back by subtracting $n$ and rebuilding the necessary bits.
The algorithmic details can be seen in \hyperref[alg:flips]{Algorithm 6}.

@@

\subsubsection{Bitmask}\label{sec:2.2.4}
The main reason, Fast-Dice-Roller is highly impractical, is due to the fact that we build up our random number bit-by-bit instead of drawing all required bits at once.
This motivates the following algorithm (see \hyperref[alg:bitmask]{Algorithm 7}) that was adopted by Apple in the macOS Sierra release when they made their own revision to the code of \texttt{arc4random\_uniform}~\cite{Apple}.

@@

%%%%% Lemire
\section{Nearly Divisionless Unbiased Sampling}\label{sec:3}
While completely avoiding division seems possible with the Fast-Dice-Roller and the Bitmask algorithm, it also comes at a cost: logarithmic time or a \texttt{clz} operation that might be even more expensive than a division on some systems.
Hence, Lemire proposed an unbiased algorithm that uses the approach of the Multiply-and-Shift algorithm coupled with rejection sampling to remove bias.
The algorithm (see \hyperref[alg:lemire]{Algorithm 8}) uses one divison operation at most and may also use no division operation at all.
It is the main result of the underlying paper of this report~\cite{Lemire}.

>>>>

\subsubsection{OpenBSD}\label{sec:2.2.1}
We somehow want to get rid of the bias generated by the last interval in \hyperref[eq:1]{equation (1)}.
The key idea is that this interval does not have to be on the right.
Instead of rejecting all values generated by a random $x \geq 2^W - (2^W \Mod n)$, we could also reject all values generated by a random $x < 2^W \Mod n$ for uniformity since the interval $[2^W \Mod n, 2^W)$ has size $2^W - (2^W \Mod n)$ which is divisible by $n$.
Due to \hyperref[lemma:1]{Lemma 1}, for every $r \in [0,n)$, we have exactly $(b - a) \div n$ numbers in $[2^W \Mod n,2^W)$ with remainder $r$, thus having a uniform distribution over $[0,n)$ when computing the remainder.
This yields \hyperref[alg:openbsd]{Algorithm 3} which stems from the C standard library in OpenBSD and macOS (\texttt{arc4random\_uniform}), and has been adopted by the Go language (\texttt{Int63n, Int31n}) with slight implementation differences~\cite{GoLang} as well as the GNU C++ standard library~\cite{GnuCpp}.

@@

\subsubsection{Java}\label{sec:2.2.2}
Since we expect remainder operations to have high latency, always needing to compute two remainders can be quite costly.
Hence, the Java language, in its \texttt{Random class}, uses a different approach.
The key insight is that when we draw a random integer $x \in [0,2^W)$ and compute its remainder $r = x \Mod n$, we can map all integers in $[0,2^W)$ to a multiple of $n$ by computing $x - r$.
Then, every number in the last interval in \hyperref[eq:1]{equation (1)} gets mapped to the last multiple $y$ of $n$ which is not greater than $2^W$.
Note that this implies $y > 2^W - n$.
Every other integer is mapped to a different (and thus smaller) multiple $z$ of $n$ for which $z \leq y - n$ and thus $z \leq 2^W - n$.
Consequently, if we reject $x$ if $x - r > 2^W - n$ and otherwise return $r$, we get a unbiased generator.
The details can be seen in \hyperref[alg:java]{Algorithm 4}.

@@

\subsubsection{Fast-Dice-Roller}\label{sec:2.2.3}
For all prior unbiased algorithms, we need to use at least one remainder operation to debias the algorithm.
Jérémie Lumbroso however devised an algorithm for sampling uniform random integers in $[0,n)$ for a given $n$ by iteratively building up the final number bit-by-bit~\cite{Flips}.
In comparison to previous algorithms, we do not use the whole $W$ bits generated by the PRNG but instead only use one bit at a time building up the final number consisting of $\lceil\log_2n\rceil + 1$ bits.
If we, at some point, overshoot and generate a number bigger than $n$ (but smaller than $2n$) we take a step back by subtracting $n$ and rebuilding the necessary bits.
The algorithmic details can be seen in \hyperref[alg:flips]{Algorithm 5}.

@@

\subsubsection{Bitmask}\label{sec:2.2.4}
The main reason, Fast-Dice-Roller is highly impractical, is due to the fact that we build up our random number bit-by-bit instead of drawing all required bits at once.
This motivates the following algorithm (see \hyperref[alg:bitmask]{Algorithm 6}) that was adopted by Apple in the macOS Sierra release when they made their own revision to the code of \texttt{arc4random\_uniform}~\cite{Apple}.

@@

%%%%% Lemire
\section{Nearly Divisionless Unbiased Sampling}\label{sec:3}
While completely avoiding division seems possible with the Fast-Dice-Roller and the Bitmask algorithm, it also comes at a cost: logarithmic time or a \texttt{clz} operation that might be even more expensive than a division on some systems.
Hence, Lemire proposed an unbiased algorithm that uses the approach of the Multiply-and-Shift algorithm coupled with rejection sampling to remove bias.
The algorithm (see \hyperref[alg:lemire]{Algorithm 7}) uses one divison operation at most and may also use no division operation at all.
It is the main result of the underlying paper of this report~\cite{Lemire}.
